\section{Regulatory Compliance Metric}

The Regulatory Compliance Metric evaluates whether an AI assistant's responses adhere to a corpus of regulatory documents. Rather than relying on an LLM judge, this metric grounds its evaluation in a retrieval-augmented pipeline: relevant passages are retrieved from the regulatory corpus and then scored against the assistant's response using a neural reranker. This design keeps evaluation transparent and auditable, producing verdicts that are directly traceable to specific regulatory passages.

\subsection{Motivation}

Deployed AI assistants in regulated domains---such as healthcare, finance, or legal services---must produce responses that align with governing regulations, policies, and compliance requirements. A mismatch between an assistant's output and a regulatory document may constitute a compliance violation, regardless of how fluent or helpful the response appears. The Regulatory Compliance Metric addresses this gap by treating the regulatory corpus as a ground truth and quantifying the degree to which each assistant response is supported or contradicted by that corpus.

\subsection{Retrieval-Augmented Evaluation Pipeline}

The evaluation pipeline proceeds in three stages: corpus ingestion, chunk retrieval, and contradiction checking.

\subsubsection{Corpus Ingestion and Chunking}

Regulatory documents are loaded via a \texttt{CorpusConnector} abstraction and chunked into fixed-size overlapping windows by the \texttt{DocumentRetriever}. Given a document of length $L$ characters, chunk size $c$, and overlap $o$, the $k$-th chunk spans characters $[k(c - o),\; k(c - o) + c]$. Consecutive whitespace normalization is applied before chunking to reduce noise. The total number of chunks across all documents constitutes the retrieval index.

All chunk texts are encoded into dense vectors by an \texttt{Embedder} instance. The \texttt{Embedder} interface exposes two encoding methods: \texttt{encode} for documents and \texttt{encode\_query} for queries, allowing implementations to apply instruction prefixes or asymmetric encoding strategies typical of bi-encoder retrieval models~\cite{reimers2019sentencebert}.

\subsubsection{Merged Chunk Retrieval}

For each interaction, the retriever issues two separate embedding lookups---one against the user query $Q$ and one against the assistant response $A$---and merges the results:

\begin{equation}
    \mathcal{C}(Q, A) = \text{deduplicate}\!\left(\text{retrieve}(Q) \cup \text{retrieve}(A)\right)
\end{equation}

where deduplication retains, for each unique (source, chunk\_index) pair, the chunk with the highest cosine similarity. This dual-query strategy ensures that chunks relevant to the question's subject matter and chunks topically aligned with the assistant's claims are both surfaced.

Cosine similarity is used as the retrieval scoring function. Let $\mathbf{q}$ be the query embedding and $\mathbf{d}_i$ the embedding of the $i$-th chunk; only chunks satisfying $\text{sim}(\mathbf{q}, \mathbf{d}_i) \geq \tau_s$ are retained, where $\tau_s$ is the configurable similarity threshold (default $\tau_s = 0.3$). The top-$k$ chunks by similarity are forwarded to the contradiction checker (default $k = 10$).

\subsubsection{Contradiction Checking via Reranking}

The retrieved chunks are re-scored by a \texttt{Reranker} that computes a relevance score $r_i \in [0, 1]$ for each chunk $\mathbf{c}_i$ with respect to the assistant response $A$. A cross-encoder architecture is typical here, as it jointly encodes the document-response pair rather than relying on independent embeddings~\cite{nogueira2019passage}.

Each chunk is assigned a binary verdict based on a configurable contradiction threshold $\tau_c$ (default $\tau_c = 0.6$):

\begin{equation}
    v_i = \begin{cases}
        \texttt{SUPPORTS}     & \text{if } r_i \geq \tau_c \\
        \texttt{CONTRADICTS}  & \text{if } r_i < \tau_c
    \end{cases}
\end{equation}

\subsection{Interaction-Level Verdict and Compliance Score}

Let $n_s$ and $n_c$ denote the number of supporting and contradicting chunks for a given interaction, respectively, and $n_t = n_s + n_c$ the total number of retrieved chunks.

If no chunks are retrieved ($n_t = 0$), the interaction is assigned an \texttt{IRRELEVANT} verdict, indicating that the regulatory corpus does not cover the topic of the query-response pair. Otherwise, the raw compliance score is:

\begin{equation}
    s = \frac{n_s}{n_t}
\end{equation}

The interaction-level verdict is then determined by comparing $s$ against a compliance threshold $\tau_p$ (default $\tau_p = 0.5$):

\begin{equation}
    \text{verdict} = \begin{cases}
        \texttt{NON\_COMPLIANT} & \text{if } n_c > 0 \text{ and } n_s = 0 \\
        \texttt{COMPLIANT}      & \text{if } s \geq \tau_p \\
        \texttt{NON\_COMPLIANT} & \text{otherwise}
    \end{cases}
\end{equation}

The first branch ensures that any interaction where every retrieved chunk contradicts the response is immediately classified as non-compliant, regardless of the threshold.

\subsection{Session-Level Aggregation}

Interaction-level compliance scores are aggregated to a session-level metric using a pluggable \texttt{StatisticalMode} interface that supports both frequentist and Bayesian paradigms. For a session with $n$ interactions producing scores $s_1, \ldots, s_n$ and weights $w_1, \ldots, w_n$:

\paragraph{Frequentist Mode.} Returns the weighted mean compliance score:

\begin{equation}
    \bar{s} = \sum_{i=1}^{n} w_i \cdot s_i
\end{equation}

\paragraph{Bayesian Mode.} Returns a bootstrapped posterior with credible intervals. Given the score vector $\mathbf{s} = (s_1, \ldots, s_n)$ and weight vector $\mathbf{w} = (w_1, \ldots, w_n)$, we draw $S$ Monte Carlo bootstrap samples (default $S = 5000$):

\begin{equation}
    \bar{s}^{(k)} = \frac{1}{n} \sum_{j=1}^{n} s_{I_j^{(k)}}, \quad I_j^{(k)} \sim \text{Categorical}(\mathbf{w}), \quad k = 1, \ldots, S
\end{equation}

The session-level compliance is then reported with a credible interval:

\begin{equation}
    \hat{s} = \mathbb{E}[\bar{s}^{(k)}], \quad \text{CI}_{1-\alpha} = \left[ Q_{\alpha/2}(\bar{s}^{(k)}), \; Q_{1-\alpha/2}(\bar{s}^{(k)}) \right]
\end{equation}

where $Q_p$ denotes the $p$-th quantile of the bootstrap distribution and $\alpha = 0.05$ for 95\% credible intervals.

The session-level verdict mirrors the interaction-level logic: if all interactions are \texttt{IRRELEVANT}, the session is \texttt{IRRELEVANT}; otherwise, the session is \texttt{COMPLIANT} if $\bar{s} \geq \tau_p$ and \texttt{NON\_COMPLIANT} otherwise.

\subsection{Data Representation}

Evaluation results are represented through a typed schema hierarchy:

\begin{enumerate}
    \item \textbf{RegulatoryChunk}: Stores a single retrieved chunk with its full provenance:
    \begin{itemize}
        \item \texttt{text}, \texttt{source}, \texttt{chunk\_index}: The chunk content and its location within the corpus
        \item \texttt{similarity}: Cosine similarity from embedding retrieval ($\in [0, 1]$)
        \item \texttt{reranker\_score}: Cross-encoder score against the assistant response ($\in [0, 1]$)
        \item \texttt{verdict}: \texttt{SUPPORTS} or \texttt{CONTRADICTS}
    \end{itemize}
    \item \textbf{RegulatoryInteraction}: Per-interaction result containing the compliance score, verdict, supporting and contradicting chunk counts, the full list of \texttt{RegulatoryChunk} records, and a human-readable \texttt{insight} string.
    \item \textbf{RegulatoryMetric}: Session-level aggregate containing:
    \begin{itemize}
        \item \texttt{session\_id} and \texttt{assistant\_id}: Session and assistant identifiers
        \item \texttt{n\_interactions}: Total number of interactions evaluated
        \item \texttt{compliance\_score}: Aggregated session compliance score $\bar{s}$
        \item \texttt{compliance\_score\_ci\_low}, \texttt{compliance\_score\_ci\_high}: Credible interval bounds (Bayesian mode only; \texttt{None} otherwise)
        \item \texttt{verdict}: Session-level \texttt{COMPLIANT}, \texttt{NON\_COMPLIANT}, or \texttt{IRRELEVANT}
        \item \texttt{total\_supporting\_chunks}, \texttt{total\_contradicting\_chunks}: Corpus-level evidence summary
        \item \texttt{interactions}: Full list of \texttt{RegulatoryInteraction} records for drill-down analysis
    \end{itemize}
\end{enumerate}

All schemas are implemented as Pydantic~\cite{pydantic2024} models with strict field validation (e.g., \texttt{compliance\_score: float = Field(ge=0, le=1)}), ensuring that scores remain within well-defined bounds throughout the evaluation pipeline.

\subsection{Architectural Considerations}

The regulatory pipeline follows the same dependency inversion and strategy patterns used throughout the framework. The \texttt{Embedder} and \texttt{Reranker} abstractions are defined as abstract base classes; concrete implementations are injected at construction time, decoupling the pipeline from any specific model provider. This allows practitioners to substitute, for example, a local sentence-transformer embedder with a cloud-hosted embedding API without modifying the evaluation logic.

The corpus is loaded lazily on the first call to \texttt{batch()}, ensuring that initialization cost is deferred until evaluation actually begins. Subsequent calls reuse the pre-built embedding index, so the $O(|D|)$ embedding cost is paid only once per \texttt{Regulatory} instance regardless of the number of sessions evaluated.
